### 2025-07-05 Text/Thesis/Chapters/Chapter4.qmd
thesis update

## ANALYSIS WITH SIMULATED DATA

@@

## APPLICATIONS TO SAR DATA

@@

>>>>

## ANALYSIS WITH SIMULATED DATA {#sec-sim2}

@@

## APPLICATIONS TO SAR DATA {#sec-sim3}

@@

## INTERACTIVE SHINY APP FOR HETEROGENEITY DETECTION

Shiny is an R package (version 1.9.1) developed by RStudio (now Posit) that makes it possible to build interactive web applications directly from R code, without requiring knowledge of HTML, CSS or JavaScript.  Users can manipulate inputs (e.g. sliders, file uploads, dropdowns) and see results update instantly in the browser.

To facilitate reproducibility and allow exploration of the heterogeneity-detection pipeline on both simulated and real SAR images, we have developed a Shiny web application.

The app supports three data‐loading modes:

- **Simulated image examples**: Choose one of two simulated images (e.g. $L=5$ or $L=9$).

- **ENVI sample images**:  Select one of three small SAR samples in ENVI format (e.g. Dublin $L=16$, London $L=1$, or New Orleans $L=12$).

- **Upload ENVI**: Upload any custom \texttt{.img}/\texttt{.hdr} pair to process your own data.

Once the image is loaded, the user can select one of three entropy estimators (Shannon, Rényi, or Tsallis), set the order $\lambda$ (for Rényi and Tsallis), specify the number of looks $L$, enable or disable bootstrap resampling (and choose the number of replicates), and fix the sliding‐window size $n \times n$. By clicking **Run Detection**, the app performs the sliding‐window test and displays the resulting $p$-value heat‐map in color.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.96\textwidth]{Figures/shiny2.png}
  \caption{Shiny application interface. The left panel shows the data‐loading and parameter controls; the right panel displays the colorized $p$-value map and elapsed runtime.}
  \label{fig:shiny-app}
\end{figure}

This interactive tool complements the results presented in Sections \ref{sec-sim2} and \ref{sec-sim3} by enabling real‐time parameter tuning and direct comparison across estimators.

The application is publicly accessible at:
[https://janeth-alpala.shinyapps.io/heterogeneity\_detection\_app/](https://janeth-alpala.shinyapps.io/heterogeneity\_detection\_app/)

### 2025-07-05 Text/Thesis/Frontmatter/abstract.tex
thesis update

This work presents a statistical approach to detect heterogeneity 
in synthetic aperture radar (SAR) intensity data using entropy-based methods.
In SAR data analysis, accurate terrain interpretation fundamentally depends 
on distinguishing two key regimes: homogeneous regions, characterized by fully 
developed speckle and modeled by the Gamma distribution, and heterogeneous areas, 
which exhibit complex scattering behaviors typically captured by the $\mathcal{G}_I^0$ distribution.
While this discrimination is essential for remote sensing applications, classical parametric tests often fail to address it 
effectively due to analytical limitations and numerical instabilities.
To overcome these challenges, we propose three test statistics for detecting heterogeneity 
in SAR imagery based on Shannon, Rényi, and Tsallis entropy measures. These tests rely on 
nonparametric entropy estimators constructed from sample spacings, avoiding explicit assumptions 
about the underlying distribution. To enhance their accuracy, particularly in small samples, 
we incorporate a bootstrap bias-correction procedure that improves the stability of the estimators, 
reduces bias, and lowers mean squared error.
The proposed tests are evaluated using Monte Carlo simulations, assessing size and power 
under varying sample sizes and numbers of looks.
Results demonstrate that the Rényi and Tsallis-based tests outperform the Shannon-based version by 
detecting more subtle texture variations while maintaining greater reliability in identifying 
truly homogeneous regions.
Finally, the methodology is applied to both simulated and real SAR data. We generate $p$-value maps using a 
sliding window analysis, allowing for visual and quantitative assessment of spatial heterogeneity. 
The Rényi-based test consistently identifies fine-scale roughness patterns, while the Tsallis-based 
test proves more effective in reliably detecting homogeneous areas. Together, these entropy-based 
tools provide a robust, interpretable, and fully unsupervised framework for heterogeneity detection 
in SAR data.

>>>>

This work presents a statistical approach to detect heterogeneity 
in synthetic aperture radar (SAR) intensity data using entropy-based methods.
In SAR data analysis, accurate terrain interpretation fundamentally depends 
on distinguishing two key regimes: homogeneous regions, characterized by fully 
developed speckle and modeled by the Gamma distribution, and heterogeneous areas, 
which exhibit complex scattering behaviors typically captured by the $\mathcal{G}_I^0$ distribution.
While this discrimination is essential for remote sensing applications, classical parametric tests often fail to address it 
effectively due to analytical limitations and numerical instabilities.
To overcome these challenges, we propose three test statistics for detecting heterogeneity 
in SAR imagery based on Shannon, Rényi, and Tsallis entropy measures. These tests rely on 
nonparametric entropy estimators constructed from sample spacings, avoiding explicit assumptions 
about the underlying distribution. To enhance their accuracy, particularly in small samples, 
we incorporate a bootstrap bias-correction procedure that improves the stability of the estimators, 
reduces bias, and lowers mean squared error.
The proposed tests are evaluated using Monte Carlo simulations, assessing size and power 
under varying sample sizes and numbers of looks.
Results demonstrate that the Rényi and Tsallis-based tests outperform the Shannon-based test by detecting subtler texture variations 
while maintaining greater reliability in identifying 
truly homogeneous regions.
Finally, the methodology is applied to both simulated and real SAR data. We generate $p$-value maps using a 
sliding window analysis, allowing for visual and quantitative assessment of spatial heterogeneity. 
The Rényi-based test consistently identifies fine-scale roughness patterns, while the Tsallis-based 
test proves more effective in reliably detecting homogeneous areas. Together, these entropy-based 
tools provide a robust, interpretable, and fully unsupervised framework for heterogeneity detection 
in SAR data.

### 2025-07-05 Text/Thesis/Frontmatter/resumo.tex
thesis update

Este trabalho apresenta uma abordagem estatística para detectar heterogeneidade em dados de 
intensidade de radar de abertura sintética (SAR) utilizando métodos baseados em entropia. 
Na análise de dados SAR, uma interpretação precisa do terreno depende fundamentalmente da 
distinção entre dois regimes principais: regiões homogêneas, caracterizadas por speckle 
totalmente desenvolvido e modeladas pela distribuição Gamma, e áreas heterogêneas, que 
exibem comportamentos de espalhamento complexos, geralmente representados pela distribuição 
$\mathcal{G}_I^0$.
Embora essa discriminação seja essencial para aplicações de sensoriamento remoto, 
testes paramétricos clássicos muitas vezes não a abordam de forma eficaz devido a 
limitações analíticas e instabilidades numéricas.
Para superar esses desafios, propomos três estatísticas de teste para detectar heterogeneidade 
em imagens SAR com base nas entropias de Shannon, Rényi e Tsallis. Esses testes utilizam 
estimadores não paramétricos de entropia construídos a partir de espaçamentos amostrais, 
evitando suposições explícitas sobre a distribuição subjacente. Para aumentar a precisão dos testes, 
especialmente em amostras pequenas, incorporamos um procedimento de correção de viés via bootstrap, 
que melhora a estabilidade dos estimadores, reduz o viés e o erro quadrático médio.
As estatísticas propostas são avaliadas por meio de simulações de Monte Carlo, 
onde analisamos seu tamanho e poder sob diferentes condições de speckle e textura. 
Os resultados mostram que os testes baseados em Rényi e Tsallis superam a versão baseada 
em Shannon, detectando variações de textura mais sutis e mantendo maior confiabilidade na 
identificação de regiões verdadeiramente homogêneas.
Por fim, a metodologia é aplicada a dados SAR simulados e reais. 
A análise é realizada com janelas deslizantes, gerando mapas de valores de $p$ que permitem a a
valiação visual e quantitativa da heterogeneidade espacial. O teste baseado na entropia de 
Rényi mostra desempenho superior na identificação de padrões de rugosidade em pequena escala, 
enquanto o teste baseado em Tsallis é eficaz na detecção de regiões homogêneas. Em conjunto, 
essas ferramentas baseadas em entropia oferecem uma estrutura robusta, interpretável e não 
supervisionada para a detecção de heterogeneidade em dados SAR.

>>>>

Este trabalho apresenta uma abordagem estatística para detectar heterogeneidade em dados de 
intensidade de radar de abertura sintética (SAR) utilizando métodos baseados em entropia. 
Na análise de dados SAR, uma interpretação precisa do terreno depende fundamentalmente da 
distinção entre dois regimes principais: regiões homogêneas, caracterizadas por speckle 
totalmente desenvolvido e modeladas pela distribuição Gamma, e áreas heterogêneas, que 
exibem comportamentos de espalhamento complexos, geralmente representados pela distribuição 
$\mathcal{G}_I^0$.
Embora essa discriminação seja essencial para aplicações de sensoriamento remoto, 
testes paramétricos clássicos muitas vezes não a abordam de forma eficaz devido a 
limitações analíticas e instabilidades numéricas.
Para superar esses desafios, propomos três estatísticas de teste para detectar heterogeneidade 
em imagens SAR com base nas entropias de Shannon, Rényi e Tsallis. Esses testes utilizam 
estimadores não paramétricos de entropia construídos a partir de espaçamentos amostrais, 
evitando suposições explícitas sobre a distribuição subjacente. Para aumentar a precisão dos testes, 
especialmente em amostras pequenas, incorporamos um procedimento de correção de viés via bootstrap, 
que melhora a estabilidade dos estimadores, reduz o viés e o erro quadrático médio.
As estatísticas propostas são avaliadas por meio de simulações de Monte Carlo, 
onde analisamos seu tamanho e poder sob diferentes condições de speckle e textura. 
Os resultados mostram que os testes baseados em Rényi e Tsallis superam a versão baseada 
em Shannon, detectando variações de textura mais sutis e mantendo maior confiabilidade na 
identificação de regiões verdadeiramente homogêneas.
Por fim, a metodologia é aplicada a dados SAR simulados e reais. 
A análise é realizada com janelas deslizantes, gerando mapas de valores de $p$ que permitem a 
avaliação visual e quantitativa da heterogeneidade espacial. O teste baseado na entropia de 
Rényi mostra desempenho superior na identificação de padrões de rugosidade em pequena escala, 
enquanto o teste baseado em Tsallis é eficaz na detecção de regiões homogêneas. Em conjunto, 
essas ferramentas baseadas em entropia oferecem uma estrutura robusta, interpretável e não 
supervisionada para a detecção de heterogeneidade em dados SAR.
